---
title: "Order Recommendation System for Instacart"
output:
  html_document:
    df_print: paged
---

#### **Instacart Overview**

##### **I ndustry Background**

Online grocery shopping is changing the whole supermarket industry. Same-day grocery delivery
industry is on trend and is predicted to worth $18 billion value by the end of 2018. Niche online
grocery services are competing over conventional grocery business with its advantages such as
convenience and selection. According to an online survey, 25% of millennials said they would pay a
premium for same-day delivery. Grocery giants such as Amazon, Costco and Target are offering
same-day grocery delivery service.

##### **What is Instacart?**

Instacart is an American company that operates as a same-day grocery on-demand delivery service
and has its headquarter at San Francisco. Instacart’s service is mainly provided through online order
by an app and Instacart personal shoppers simply go to a store and purchase the ordered items at a
chosen retail. As of March 2017, Instacart services 36 markets, composed of 1,200 cities in 25
states including California, Arizona and New York. In the same month it was valued at
approximately $3.4 billion dollars and was considered “one of the most valuable companies in the
America.”

##### **I nstacart’s Business Model**

Instacart Customers are given a decision to shop from any store that is situated nearby. Personal
shoppers who are found nearby the store will receive the order and starts collecting items as
mentioned in the order. Shoppers are paid hourly rate while customers are also allowed to tip the
shoppers. Shoppers pays the bill through Instacart’s prepaid debit card which is accepted at the
store. The shopper then goes to deliver the groceries to the customer.

##### **I nstacart’s Revenue Model**

Delivery charges: Delivery charges are Instacart’s fundamental profits. The charges may
be higher if the customer requires speedier delivery or in busier circumstances. There are
two options: Order above $35 entices a fixed delivery charge of $3.99 to $5.99; orders
less than $35 are priced $7.99 to $9.99.

  • Membership fee: Instacart also extends an annual membership scheme priced at $99 and
is termed “Instacart Express”. Users can opt for the membership to get groceries delivered
at no cost for the entire one year.
  • Mark-up charges: Products sold through the Instacart platform, some are priced as the same
as store while in some cases priced 15% higher than the retail price. The revenue generated
through these mark-up fees goes to the Instacart wallet.
  • First Free Delivery: Instacart hypnotizes customers with a free delivery options for users
who chooses to use the Instacart facility app for the first time, so pretty much a smart way
to solve the “cold start” problem and provide more incentive for new app users to place an
order.

##### **Dataset**

The dataset published by Instacart itself, called “The Instacart Online Grocery Shopping Dataset
2017”, is a relational set of files describing customers’ orders over time. The dataset is anonymized
and contains a sample of over 3 million grocery orders from more than 200,000 Instacart users. For
each user, between 4 and 100 of their orders are provided, with the sequence of products
purchased in each order.

The “aisles” table (Figure 1 left) shows all aisles and their unique aisle id.
The “department” table (Figure 1 right) shows the departments and their unique department
id.

The “products” table (Figure 2) shows the products and the aisles and departments they belong
to. It can be joined with aisles and departments table via aisle_id and department_id.

The “order_products_” tables contains two sub-tables, which are order_products_prior table
and order_products_train table. They are the most essential tables we are using for our analysis.
They specify which products were purchased in each order and the add-to-cart order of each
product purchased. They also contain a binary attribute ‘reordered’, with 1 indicates the customer
has a previous order that contains the product. These tables can be connected with products table
via product_id and with orders table via order_id.

The “orders” table, which contains 3,421,083 purchasing records, tells to which evaluation set
(prior, train, test) an order belongs. Plus the day and time an order was placed, which may help us
determine when is the best time to send the recommendation. It is another essential table for our
analysis.

##### **Questions**

Customer satisfaction is a major influence factor of customer retention rate. Most of Instacart users
are busy people who do not have time going to grocery store on their own. By predicting which
products an user will buy again and recommend these products to users during their shopping
process, Instacart may offer better convenience to its users. Therefore, customer buying activities
and satisfaction rate are both likely to increase, which not only drives to a higher revenue, but also
more loyal customers.

In this project, the core question is: Can we increase active customers and customer buying
activities by improving the order recommendation system based on their previous orders?
To know when and what should we recommend to each customer, we need to glean some insights
on customers’ order placing behaviors. Therefore, we can break down the core question into the
following sub-questions:

1) When should we send our recommendation?

2) How many items should our recommendation contain?

3) What products should we recommend?

4) Bonus: output your prediction on test dataset and show which kind of products would mostly
reordered.

*prior- purchase order in the past
*train- the very last purchase
*order_number: how many times this customer buy the product

##### **Load Data**
 
```{r}
rm(list = ls())
library(dplyr)
library(ggplot2)
library(gridExtra)


aisles <- read.csv("dataset/aisles.csv")
departments <- read.csv("dataset/departments.csv")
order_products__prior <- read.csv("dataset/order_products__prior.csv")
order_products__train <- read.csv("dataset/order_products__train.csv")
orders <- read.csv("dataset/orders.csv")
products <- read.csv("dataset/products.csv")


```

##### **Data Overview**
##### **aisles**
```{r}

head(aisles)
str(aisles)
summary(aisles)


```

##### **Observation: 134 types of aisles**

##### **departments**

```{r}

head(departments)
str(departments)
summary(departments)

```
##### **Observation: 21 department**


##### **products**


```{r}
head(products)
str(products)
summary(products)
```

##### **Observation: 49688 products**



##### **order_products__prior**
```{r}

head(order_products__train)
str(order_products__prior)
summary(order_products__train)

```

##### **Observation: 843370 orders, with max 80 add to cart order.**



##### **orders**
```{r}

head(orders)
str(orders)
summary(orders)
```
##### **Observation **

'order_dow' means 'order day of week', 0 stands for monday, it is not missing value
'order hour of day', 0 stand for the first hour of a day which is 1 am
206209 NAs' for days since prior order in table orders which count 5% of total
evaluation set for train has 131209 sets and test for 75000 sets which we predict reordered items only for the test data set


### Exploratory Data Analysis (EDA) ### 

#####1) when do customer order


```{r}

#str(orders)
# orders and hours distribution 

p1<-orders%>% ggplot(aes(x = order_hour_of_day)) + geom_bar(stat = 'count',fill = "orange") + ylab('number of orders')+
  scale_y_continuous(labels = scales::comma)+ggtitle('Total Number of Orders in Hour of Day') +
  theme(plot.title = element_text(hjust = 0.5))
p1

```

####Observation:

**Most orders are from 8am to 18pm daily**


```{r}
#orders and weekdays distribution

p2 <- orders %>% ggplot(aes(x = order_dow)) + 
  geom_bar(stat = "count", fill = "orange") + ylab('number of orders') + 
  scale_y_continuous(labels = scales::comma) +
  ggtitle('Number of Orders in Day of Week') +
  theme(plot.title = element_text(hjust = 0.5))
p2

```


**Most order are amongst monday and Tuesday**

```{r}

# days since prior order

p3 <- orders %>% filter(days_since_prior_order != '')%>%
  ggplot(aes(x = days_since_prior_order))+
  geom_bar(stat = "count", fill = "orange") + ylab('Number of orders') + 
  xlab('Days since prior day')+
  scale_y_continuous(labels = scales::comma) +
  ggtitle('Days since Prior Order') +
  theme(plot.title = element_text(hjust = 0.5))
p3

```

**Reorder frequency shows increase trend in first 7 days, then decreases, the highest frequency of re-order is once per month and then once per once per week **


##### Part 2: how many prior orders


```{r}

p4 <- orders%>%filter(eval_set == 'prior') %>% count(order_number) %>% 
  ggplot(aes(order_number,n)) + geom_point(color = "blue", size = 2) +
  ggtitle('Number of Prior Orders of Customers') + 
  theme(plot.title = element_text(hjust = 0.5))+
  ylab('number of customers') + xlab('nubmer of prior orders')

p4


```



**the dataset include users that have 4~100 orders, we can see that there are at least 3 prior orders for each consumer**
**the number of customer is decreasing as the order_number increase**



#####5 how many items each customer buy each time?



```{r}

summary(order_products__prior)


```


```{r}

##prior table

p51 <- order_products__prior %>% group_by(order_id) %>%
  summarise(number_of_items = last(add_to_cart_order)) %>%
  ggplot(aes(x = number_of_items)) + geom_bar(stat = 'count',fill='blue') + xlim(0,80) + 
  ggtitle('Number of Items of each Order in prior table')+
  theme(plot.title = element_text(hjust = 0.5)) + ylab('number of orders')


## train table
p52 <- order_products__train %>% group_by(order_id) %>% 
  summarise(number_of_items = last(add_to_cart_order)) %>%
  ggplot(aes(x = number_of_items)) +
  geom_histogram(stat = 'count', fill = 'orange') + xlim(0,80) +
  ggtitle('Number of Items of Each Order in train table') +
  theme(plot.title = element_text(hjust = 0.5)) + ylab('number of orders')
grid.arrange(p51,p52,nrow=2)

```


######How many of the items are reordered?

```{r}
df6<- order_products__prior%>%group_by(reordered)%>%
  summarise(count = n()) %>%
  mutate(proportion = round(count/sum(count),2))
df6


```


```{r}

p6<-df6%>%ggplot(aes(x=reordered,y = count,fill = reordered))+geom_bar(stat = 'identity')+ theme(legend.position = "none") + 
  scale_y_continuous(labels = scales::comma) + ylab('number of items') +
  ggtitle('Number of Reordered Items') + # geom_col(aes(fill = reordered)) 
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_text(aes(label = proportion)) 
p6 

```


**59% of the items in prior table arfe re-ordered**



##Part 3: What are the popular products?

```{r}
#add to cart order : the order of the item placed in the cart
df7 <- order_products__prior %>% group_by(product_id)%>% summarise(proportion_reordered = mean(reordered),num_of_items = n())
df7


```



```{r}




df71 <- df7 %>% arrange(desc(num_of_items))%>% left_join(products, by = 'product_id')%>% head(10)

p7 <- df71 %>% ggplot(aes(x=reorder(product_name, -num_of_items), y=num_of_items))+
  geom_bar(stat = 'identity', fill = "green") + scale_y_continuous(labels = scales::comma) +ylab('number of items') + theme(axis.text.x = element_text(angle = 90,hjust = 1),axis.title.x = element_blank()) + ggtitle('The Top 10 mostly Purchased Products') +  theme(plot.title = element_text(hjust = 0.5))
p7
```
**Banana is the best seller**

##### what product in their first choice add to cart?

```{r}

df8<-order_products__train%>% group_by(product_id, add_to_cart_order) %>%
  summarise(count = n()) %>% mutate(pct = count/sum(count))
#df8

df81 <- df8 %>% filter(add_to_cart_order ==1, count>20) %>%left_join(products,by = 'product_id')%>%
  select(product_name,pct,count) %>% arrange(desc(pct))%>%head(10)

#df81

p8 <- df81 %>% ggplot(aes(x=reorder(product_name,-pct), y=pct))+
  geom_bar(stat="identity",fill="green")+
  theme(axis.text.x=element_text(angle=90, hjust=1),axis.title.x = element_blank()) + 
  coord_cartesian(ylim=c(0.4,0.7)) +
  ggtitle('The Top 10 Products that are first added to cart') +
  theme(plot.title = element_text(hjust = 0.5))
p8
```


**white multifold towels have the highest percentage add to cart**


#####  Organic food vs non-organic food


```{r}
library(stringr)

df9<- products%>% mutate(organic = ifelse(str_detect(str_to_lower(products$product_name),'organic'),'organic','not organic'),organic = as.factor(organic))
head(df9)
prop.table(table(df9$organic))

```

**10% products are organic while 90% products are not organic**


#####how is the re-ordered percentage of organic and non-organic product

```{r}
df10 = order_products__train %>% left_join(df9, by = 'product_id') %>% group_by(organic) %>%
  summarise(mean_reordered = mean(reordered))
df10 %>% ggplot(aes(x= organic,y=mean_reordered, fill = organic)) + 
  geom_bar(stat = 'identity') +
  geom_text(aes(label = round(mean_reordered, 2), y= mean_reordered + 0.03)) +
  ggtitle('Reordered Percentage vs Organic') +
  theme(plot.title = element_text(hjust = 0.5))
```

####organic product has 65 percentages reordered,although organic product only counts 10 percent as total but reorder rate of organic product is much higher than non-organic 


#####Part 4: visualizing the product portfolio

#####How are aisles organized within departments?




```{r}
#install.packages('treemap')
library(treemap)
df11 <- products %>% group_by(department_id, aisle_id) %>% summarize(n=n()) %>%
  left_join(departments,by="department_id") %>%
  left_join(aisles,by="aisle_id")
df11
map1 <- order_products__prior %>% 
  group_by(product_id) %>% 
  summarize(count=n()) %>% 
  left_join(products,by="product_id") %>% 
  ungroup() %>% 
  group_by(department_id,aisle_id) %>% 
  summarize(sumcount = sum(count)) %>% 
  left_join(df11, by = c("department_id", "aisle_id")) %>% 
  mutate(onesize = 1)
map1


# as we can see n means how many products in one deparment one aisle.
# sumcount means how many time the product has been ordered
```

```{r}
treemap(map1,index=c("department","aisle"),
        vSize="onesize",vColor="department",
        palette="Set3",title="Aisles organized within Departments",
        sortID="-sumcount", 
        border.col="#FFFFFF",type="categorical", 
        fontsize.legend = 0,bg.labels = "#FFFFFF")
```


#####How many unique products are offered in each department/aisle?

```{r}
treemap(df11,index=c("department","aisle"),
        vSize="n",title="Aisles (affected by unique products number) organized within Departments",
        palette="Set3",border.col="#FFFFFF")


```


#####How often are products from the department/aisle sold?
```{r}
treemap(map1,index=c("department","aisle"),
        vSize="sumcount",title="Aisles (affected by products quantity sold) organized within Departments",
        palette="Set3",border.col="#FFFFFF")
```


##### Main take-aways
 1. When to recommend:
 - Recommend at 7 days and 30 days after previous order.
 - Recommend at 8:45 or 1:00 when users at break.
 - Recommend when there are discounts in markets.

 2. How many items to recommend:
 - Mostly 4 to 7 items at a time, maximum 10.
 - Divided into sublists for user-friendly viewing.

 3. What to recommend:
 - Popular products based on chosen aisles and departments of users.
 - Related or potentially-interested aisles based on his ordered items.



























